---
title: "Mollified Two-Piece Distributions"
author: "[F Javier Rubio](https://sites.google.com/site/fjavierrubio67/)"
date: "25/06/2017"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Mollified Two-Piece Distributions

[Mollifiers](https://en.wikipedia.org/wiki/Mollifier) can be used to smooth a density function. In particular, "mollifying" a two-piece density (or a double-two-piece density) function produces a [smooth](https://en.wikipedia.org/wiki/Smoothness) density which can be made arbitrarily close to the original two-piece density by controlling the scale parameter of the Mollifier. The "mollification" process removes the unpleasant lack of twice-differentiability at the mode in the family of two-piece distributions (the mollified density has derivatives of all orders). "There is no such thing as a free lunch": "mollified" densities are more difficult to evaluate. The following R code shows how to "mollify" some members of the two-piece family (see also 'twopiece', 'TPSAS', and 'DTP' [R packages](https://sites.google.com/site/fjavierrubio67/resources)). 

**References.**

1. [Mollifiers. From R. Showalter’s book on PDEs](https://www.math.psu.edu/~anovikov/acm105/mollifiers.pdf).

2. [Inference in Two-Piece Location-Scale Models with Jeffreys Priors](http://projecteuclid.org/euclid.ba/1393251764).

3. [Bayesian modelling of skewness and kurtosis with Two-Piece Scale and shape distributions](http://projecteuclid.org/euclid.ejs/1440680330).

4. [On modelling asymmetric data using two-piece sinh–arcsinh distributions](http://projecteuclid.org/euclid.bjps/1469807223).

```{r}
rm(list=ls())
# Required packages
library(twopiece)
library(DTP)
library(TPSAS)

##################################################################################
# Mollifier (described in the Wikipedia entry)
##################################################################################

eta = 0.01 # Scale of the Mollifier

# Temporary function to obtain the normalising constant of the Mollifier
tempf0 = Vectorize(function(x) ifelse(abs(x/eta)<1, exp(-1/(1-(x/eta)^2))/eta,0) )
# Normalising constant
cons = integrate(tempf0,-eta,eta)$value

# The Mollifier
phi = Vectorize(function(x) ifelse(abs(x/eta)<1, exp(-1/(1-(x/eta)^2))/(eta*cons),0) )

integrate(phi,-eta,eta) # It integrates 1
# Plot of the Mollifier
curve(phi,-eta,eta,n=1000,xlab=~epsilon,ylab="Mollifier",cex.axis=1.5,cex.lab=1.5)
box()
##################################################################################
# Mollified two-piece normal: mu=0, sigma=1, gamma=0.5, "eps" parameterisation
##################################################################################

# Original two-piece normal density
dtpn = Vectorize(function(x) dtp3(x,0,1,0.5,FUN=dnorm,param="eps"))

# Mollified two-piece normal density
dmtpn = Vectorize( function(x){
  tempf = Vectorize(function(t) dtpn(x-t)*phi(t))
  val = integrate(tempf,-eta,eta)
  return(val$value)
}
)

# Comparison: virtually equal (running time: ~5 seconds)
curve(dtpn,-5,3,n=500,lwd=2,xlab="x",ylab="Density",cex.axis=1.5,cex.lab=1.5)
curve(dmtpn,-5,3,col="red",n=500,lwd=2,lty=2,add=T)
box()
legend(-4.5, 0.35, c("TP","Mollified TP"),
       text.col = "black", col= c("black","red"), lwd=2, lty = c(1, 2),
       merge = TRUE, bg = "gray90")

##################################################################################
# Mollified two-piece Laplace: mu=0, sigma=1, gamma=0.5, "eps" parameterisation
##################################################################################

# Laplace density
dlaplace <- function(y, mu=0, sigma=1, log=FALSE){
  logPDF  <- (-log(2) - log(sigma) - abs(y-mu)/sigma)
  ifelse(is.numeric(logPDF), ifelse(log, return(logPDF), return(exp(logPDF))), 
         logPDF)
}

# Original two-piece Laplace density
dtplap = Vectorize(function(x) dtp3(x,0,1,0.5,FUN=dlaplace,param="eps"))

# Mollified two-piece Laplace density
dmtplap = Vectorize( function(x){
  tempf = Vectorize(function(t) dtplap(x-t)*phi(t))
  val = integrate(tempf,-eta,eta)
  return(val$value)
}
)

# Comparison: virtually equal (running time: ~5 seconds)
curve(dtplap,-7.5,3,n=500,lwd=2,xlab="x",ylab="Density",cex.axis=1.5,cex.lab=1.5)
curve(dmtplap,-7.5,3,col="red",n=500,lwd=2,lty=2,add=T)
box()
legend(-6.5, 0.5, c("TP","Mollified TP"),
       text.col = "black", col= c("black","red"), lwd=2, lty = c(1, 2),
       merge = TRUE, bg = "gray90")

##################################################################################################
# Mollified two-piece sinh-arcsinh: mu=0, sigma=1, gamma=0.5, delta=0.75, "eps" parameterisation
##################################################################################################

# Original two-piece sinh-arcsinh density
dtpsas0 = Vectorize(function(x) dtpsas(x,0,1,0.5,0.75,param="eps"))

# Mollified two-piece sinh-arcsinh density
dmtpsas = Vectorize( function(x){
  tempf = Vectorize(function(t) dtpsas0(x-t)*phi(t))
  val = integrate(tempf,-eta,eta)
  return(val$value)
}
)

# Comparison: virtually equal (running time: ~5 seconds)
curve(dtpsas0,-7.5,3,n=500,lwd=2,xlab="x",ylab="Density",cex.axis=1.5,cex.lab=1.5)
curve(dmtpsas,-7.5,3,col="red",n=500,lwd=2,lty=2,add=T)
box()
legend(-7, 0.3, c("TP","Mollified TP"),
       text.col = "black", col= c("black","red"), lwd=2, lty = c(1, 2),
       merge = TRUE, bg = "gray90")

######################################################################################################################
# Mollified double two-piece t: mu=0, sigma=1, gamma=0.5, delta1=1, delta2=5, "eps" parameterisation
######################################################################################################################

# Original double two-piece t density
ddtpt = Vectorize(function(x) ddtp(x,0,1,0.5,1,5,f=dt,param="eps"))

# Mollified two-piece t density
dmdtpt = Vectorize( function(x){
  tempf = Vectorize(function(t) ddtpt(x-t)*phi(t))
  val = integrate(tempf,-eta,eta)
  return(val$value)
}
)

# Comparison: virtually equal (running time: ~5 seconds)
curve(ddtpt,-10,3,n=500,lwd=2,xlab="x",ylab="Density",cex.axis=1.5,cex.lab=1.5)
curve(dmdtpt,-10,3,col="red",n=500,lwd=2,lty=2,add=T)
box()
legend(-9, 0.3, c("TP","Mollified TP"),
       text.col = "black", col= c("black","red"), lwd=2, lty = c(1, 2),
       merge = TRUE, bg = "gray90")
```